Scaling of Electrode-Electrolyte Interface Model Parameters In Phosphate Buffered Saline

### 2014-04-17 paper.tex
Ready for submission

\begin{abstract}
We report how the impedance presented by a platinum electrode scales with the concentration of phosphate buffered saline (PBS).
{
    \color{blue}
    We measure this response in various solutions with an electrode array as is commonly used with spinal cord stimulation (SCS) implants. We match the parameters of a compact non-linear electrode/electrolyte interface model to measured responses at various concentrations of PBS.
}
We find that the constant phase element of the model scales with approximately the log of concentration, whereas the resistivity is inversely proportional.
Using a novel DC measurement technique we show that the Faradaic response of a platinum electrode, and thus the safe exposure limit, does not scale with concentration below 900\thinspace mV overpotential across a pair of electrodes.
We compare objective measurements made in saline to those made in the spinal cavity of live sheep. We comment upon the appropriateness of using PBS as a substitute for in-vivo measurements.
\end{abstract}

\begin{IEEEkeywords}
    Electrical stimulation, Bioelectric phenomena, Biophysics, Bioimpedance, Biomedical electrodes, Biomedical measurements, Implantable biomedical devices
\end{IEEEkeywords}

@@

Having a compact model of an electrode in the appropriate electrolyte allows circuit designers to simulate their designs with valid loads. Knowing the effect upon the model of changes in the electrode or electrolyte would allow designers to anticipate circumstances that will affect the load seen by their circuits. Another appeal of a compact model stems from the fact that electrode characteristics are succinctly and objectively represented by the model parameters. These enable direct comparison of different electrodes, or the change in electrode properties over time. For example, in \cite{Kane13} changes in chronically-implanted electrodes are observed but there is no standard quantitative way of presenting the changes. Similarly, while there is a good understanding of what represents safe exposure, a compact model with parameters permits the prediction of the safety of any given stimulus regime.~\cite{Merrill05}

@@

Electrodes in the laboratory are typically tested in a saline solution selected to mimic the circumstances in which they will operate when implanted. In an implanted setting there is no reference electrode and many of the well established electrochemical measurement regimes do not apply. Here we measure the behaviour an electrode/electrolyte situation as it would be seen from an implant device.

The electrode used in this work is a commercial linear array of eight platinum electrodes intended for SCS implantation called an ``Octrode''.~\cite{StJudeOctrode} A picture of an Octrode is presented in figure~\ref{fig:octrode}. A solution of Phosphate Buffered Saline (PBS) diluted to one-tenth concentration (0.1X) is common in the case of Spinal Cord Stimulators (SCS), while full concentration (1.0X) PBS is considered more representative of the situation in blood.

% {
%     \color{red}
% In this manuscript we report how the parameters of a non-linear model change as the concentration of saline is varied. We refine a series of measurements compatible with commercially available medical electrodes in physiological saline that allow us to describe the impedance with minimal parameters. We place these parameters in context with a biological measurement made in the spinal cavity of a live sheep.

%     THIS IS IN THE ABSTRACT
% }

Six solutions ranging from 1.0X to 0.025X the concentration of a stock PBS solution (by mass) have been examined.
The ingredients of that stock solution are given in table~\ref{tab:PBSrecipe}.

@@

    \begin{figure}
        \begin{center}
            \includegraphics[width=230pt]{graphics/interfaceSchematic_noMemristive}
        \end{center}
        \caption{Electrical schematic of the electrode-interface impedance model used in this work. $D_{a}$ and $D_{b}$ represent diodes; $CPE$ and $R_{s}$ represent the constant phase element and series resistance respectively.}
        \label{fig:schematic}
    \end{figure}

    In this manuscript we use an adapted version of the compact non-linear model presented in \cite{ScottSingle2013}. The model is compact in that it is suitable for entering into SPICE, i.e., freely available electrical circuit simulation software. The schematic for the model used here is presented in Figure \ref{fig:schematic}.
}

@@

It has been suggested that the primary coupling mechanism between platinum electrodes and physiological saline is the reversible evolution of gas, probably oxygen, at the platinum surface, which dissolves without bubble formation.\cite{Greatbatch1969} Understanding the exact mechanism of displacement or Faradaic charge transfer at the electrode surface is beyond the scope of this work. Rather, we wish to characterise the electrode/electrolyte system in terms of a {\color{blue} small number of} parameters that may be fed into an electrical model.

{
    \color{blue}
    The remainder of this manuscript is broken into five sections. Section \ref{sect:resistorMesh} details the measurement of the inter-electrode resistances and the generation of a resistor network to replicate these resistances.
    Sections \ref{sect:cpe} and \ref{sect:faradaic} are concerned with the measurement and parameter fitting of displacement currents and Faradaic currents, respectively, across the electrolyte interface. Section \ref{sect:in-vivo} compares the results obtained in PBS with measurements in-vivo. Finally, in Section \ref{sect:conclusion} we summarise our findings and give concluding comments.
}

@@

The electrical impedance between two electrodes in an electrolyte arises from two interface impedances in series with the resistance due to the bulk of the electrolyte itself. To quantify the impedance presented by a single interface we first need to determine inter-electrode resistance presented by the electrolyte bulk. {\color{blue} Finding these resistances allows us to quantify the interface's impedance via subtraction.} As there are always two interface impedances between any two electrodes, inter-electrode resistances cannot be measured in isolation.

In \cite{ScottSingle2013}, a series of transresistance measurements were used in conjunction with a physical description of the electrode geometry to build up a representative network of resistors. That resistor network was defined using five parameters: $R_{eri}$, $R_{lri}$, $R_{li}$, mesh depth, and edge depth. Readers are referred to \cite{ScottSingle2013} for the interpretation of these model parameters. Repeating that work, we have created a network of resistors that mimic the resistances due to the solution's bulk conductivity.

The mesh depth and padding values of five columns and three rows, respectively, has been taken directly from \cite{ScottSingle2013}. {\color{blue} These parameters, together with an array of 8 electrodes, leads to a generated network of 205 resistors, the layout of which is presented as figure \ref{fig:mesh}.}

A numerical fit is then made to the two independent scaling parameters $R_{eri}$ and $R_{li}$
\footnote{The values of $R_{eri}$ and $R_{sri}$ are defined by the lengths of the electrode and inter-electrode spacers respectively as described in \cite{ScottSingle2013}. The two parameters are proportional to each other by the ratio of their lengths. Hence, we define $R_{sri}$ to be 3/4 that of $R_{eri}$ making it a dependant parameter.}.
The resulting fit and parameter values are presented in figure~\ref{fig:transresistance} and table~\ref{tab:RESparams}.

Measurements of the inter-electrode impedance are made using a Tektronix TPS2014 DSO with four fully isolated and floating channels; an Agilent 33220A Function Generator; and a desktop PC {\color{blue} running Python 2.7 for instrumentation control and result processing.} The pH level of the stock solution is measured using a calibrated EDT Instruments BA-350 pH meter. It has a pH of 7.4 and no pH adjustments were made to the derived solutions.

{
    \color{blue}
    To be clear, the resistive mesh created in this step models \textit{only} the resistivity of the solution bulk. No contribution from the interface itself appears in these measurements. The is because voltages are always measured across pairs of non-driven electrodes and the measurements are of suitably high impedance. We determine the resistive component of the interface, $R_{S}$, in the next section.
}

@@

\section{Displacement Parameters {\color{blue}/ CPE}}
\label{sect:cpe}
Displacement currents are responsible for capacitive behaviour at interfacial boundaries.
They are brought about by the redistribution of charge in response to applied fields, i.e. a charged electrode will attract or repel $Cl^{-}$ from its surface and reorientate polar molecules in the surrounding solution.\cite{Merrill05} This capacitive behaviour may also be the result of electrode polarisation in the form of reversible Faradaic reactions at the surface of the electrode.
Such reactions involving water and platinum, as identified in \cite{Horch2004,Mohtashami2011,Merrill05}, include:
\begin{eqnarray}
    Pt + H_{2}O &\Leftrightarrow& PtO + 2 H^{+} + 2 e^{-}\\
    PtO + H_{2}O &\Leftrightarrow& PtO_{2} + 2 H^{+} + 2e^{-}\\
    Pt + H^{+} & \Leftrightarrow & Pt-H\\
    Pt + H_{2}O + e^{-} &\Leftrightarrow& Pt-H+OH^{-}
\end{eqnarray}

Displacement currents are capable of transferring charge between electrical and biological systems without damaging electrodes or tissue.\cite{Horch2004}
This behaviour is modelled by way of a Constant Phase Element (CPE), also known as a fractional capacitor.
For simulation purposes this element is realised as an array of R-C branches.~\cite{ScottSingle2013,Morrison59,Elwakil10} where the value of each R-C pair is chosen to set the slope of impedance relative to frequency. The placement of R-C pairs within the CPE is determined by Morrison's parameters $m$ and $k$, where $k$ determines the spacing, or density, of R-C branches and $m$ determines the resulting slope. {\color{blue} The R-C pairs were generated via Python script to span 1\thinspace nHz--1\thinspace GHz with a spacing of 3 pairs per decade. The final CPE contains 55 branches i.e. 110 elements are used to create the CPE model.}

We measure displacement currents by sweeping the frequency of a sinusoidal stimulus across electrodes 2 and 7 of the Octrode while measuring the voltage between electrodes 2 and 3 (numbering shown in Fig.~\ref{fig:octrode}).
This allows for the measurement of a single interface impedance in series with the resistance presented by the bulk solution.

\begin{figure}
    \begin{center}
        \includegraphics{graphics/chargeInjectionEffectiveVsFrequency_magnitude}
    \end{center}
    \caption{\color{blue} Per phase charge injected during CPE measurement in each concentrations of PBS, calculated using an effective surface area of 14\thinspace mm$^{2}$. {\color{red} IS THIS CHEATING? (I have given the charged injected in terms of effective surface area instead of geometric)}}
    \label{fig:chargeInjectionVsFrequency}
\end{figure}

{
    \color{blue}
    Current density during measurements, calculated using an effective surface area of $14\thinspace mm^{2}$, peaked at $161\thinspace \mu A/cm^{2}$ at 10\thinspace kHz in 1.0X PBS and fell as low as $197\thinspace nA/cm^{2}$ at 50\thinspace mHz in 0.025X PBS.
}Charge injection per phase is shown in figure~\ref{fig:chargeInjectionVsFrequency} for each solution.
The stimulus waveform was set at 300\thinspace mV (peak) at each measurement point. The measurement instruments used are the same as those used to measure the inter-electrode resistance.
Figure~\ref{fig:CPE_Magnitude} shows the magnitude of the measured response of each of the six PBS concentrations accompanied by simulation results. Figure~\ref{fig:CPE_Phase} shows the phase response for the same measurements, again with the simulated response at each concentration. The simulated responses are generated using the optimised parameter values presented in Table~\ref{tab:CPEparams}. Those values were chosen as they give the closest match to the measured data.

@@

Figure~\ref{fig:CPE_Scaling} shows measured data and the corresponding fits for both the CPE magnitude at 0.05\thinspace Hz and $R_{S+bulk}$. The model parameters $m$, $k$, $|Z|\thinspace @\thinspace 1Hz$, and $R_S$ are shown in table~\ref{tab:CPEparams}, with $|Z|\thinspace @\thinspace 1Hz$, and $R_S$ given as functions of concentration.
Here we express the CPE offset parameter as $|Z|\thinspace @ \thinspace 1Hz$, but the fits used $|Z|\thinspace @ \thinspace 0.05Hz$, as data points at this frequency are outside the transitional zone where the impedance of the CPE and that of $R_{S+bulk}$ overlap.

The data presented in Fig.~\ref{fig:CPE_Magnitude} show a large, linear, relative shift of $|Z|$ with salinity at higher frequencies, where the resistive component of the solution dominates.
Conversely, at lower frequencies where the $|Z|$ changes with frequency, the relative change is closer to logarithmic. This region is modeled by the CPE.
This suggests that the CPE does not chiefly rely on the added salt ions, but they do have some effect.
Instead, the constant-phase charge effect must arise from phenomena not associated with the salt ions.

@@

\begin{figure}
    \begin{center}
        \includegraphics{graphics/scalingFactors_Displacement_IEEE}
    \end{center}
    \caption{Simulated and measured values of CPE parameters versus concentration of PBS. $|Z|$ is compared at 0.05\thinspace Hz as measured data is unaffected by $R_{S}$ at this frequency.}
    \label{fig:CPE_Scaling}
\end{figure}

\begin{table}
    \caption{Displacement parameter scaling. $conc.$ refers to the mass concentration of PBS.}
    \label{tab:CPEparams}
    \begin{center}
        \begin{tabular}{r | l}
            Parameter & Value \\
            \hline
            $m$ & $1.34$ \\
            $k$ & $1.773$\\
            $|Z|\: @\: 1Hz$ $(\Omega)$& $3284 \times conc^{-0.158}$ \\
            $R_{S}$ $(\Omega)$& $13.38 \times conc^{-0.8397} $\\
        \end{tabular}
    \end{center}
\end{table}

@@

Our model uses two reverse-connected diodes to represent Faradaic current conduction between electrode and electrolyte. Specifically we use the diodes to model irreversible Faradaic reactions. (Any reversible reactions, those producing reactants bound to the surface of the electrode, are encapsulated by the displacement branch (CPE) of the model.)
Possible irreversible Faradaic reactions for platinum electrodes {\color{blue}(each with a different reaction potential)} in saline are:

@@

We measure Faradaic conduction by separating it from the displacement behaviour of the CPE. By using a DC measurement technique, as opposed to cyclic measurements, we are able to separate the effects in time.
Like a conventional capacitor, when the voltage across the CPE changes, it responds by absorbing or ejecting charge, causing a spike in current. The CPE will draw negligible current once it has settled and in the case of our model we assume the remaining current conduction to be the result of Faradaic processes.

@@

Measurements commenced with a 4.5 litre solution of 1.0X PBS which was progressively diluted in factors of two between each measurement. The solution was mixed continuously using a standard laboratory grade magnetic stirrer. The solution was in equilibrium with air prior to and during measurement and the electrodes remained submerged at all times.
A single measurement run involved stepping the voltage across the electrodes from 0.5\thinspace V to 1.2\thinspace V in increments of 0.05\thinspace V; this range was previously determined to capture the onset of Faradaic conduction.

@@

In \cite{Cogan2008} it has been noted that results obtained using cyclic voltammetry are dependant on, among other factors, the measurement sweep rate. This indicates a lack of isolation between displacement and Faradaic mechanisms. Our measurements were repeated using wait-times of 4, 16, 32 and 64 seconds between voltage increments, allowing us to determine the effect of sweep rate. From these measurements and those of a separate investigation\footnote{We took measurements of the interface's response to voltage steps when left to settle for 10 000 seconds, of which the first 250\thinspace seconds are shown in figure~\ref{fig:CPE_currentVsTime}. Those measurements show a trend of decreasing current stabilisation time as the overpotential steps increase. Sixty four seconds after stepping from 0.64\thinspace V to 0.72\thinspace V, the current has stabilised to $95\%$ of its value at 10\,000 seconds. At overpotential steps for higher voltages, the settling time is further reduced. These measurements were made in still solution and we expect stirring will further reduce settling time.} we concluded that a wait-time of 64 seconds was sufficient.

@@

In figure~\ref{fig:StepResponse_Faradaic}, electrical current measurements are plotted against time (solid lines). Each spike marks the point at which the voltage is incremented by 50\thinspace mV. Note that the amount of charge absorbed by the CPE increases with concentration of PBS, but the current for each concentration converges to approximately the same value. Dotted lines have been added that link the electrical current measurements taken 10 seconds after each increment for 0.125X PBS (lower trace) and 1.0X PBS (upper trace). These dotted traces show results typical of cyclic methods, where the electrode overpotential is never constant. This illustrates the benefit of using DC measurements when measuring Faradaic currents.

Figure~\ref{fig:faradaic_logCurrentVsVoltage} shows the settled currents plotted on a log scale, versus the voltage applied across the electrodes. This uses the same data as the previous figure but has been processed so that each point represents the average of the final 24 current measurements at each step. Below 0.9\thinspace V it appears that the concentration of the solution had no identifiable effect on Faradaic conduction as there is overlap between each trace and the mean values are not monotonic with concentration in this region.  Between 0.9\thinspace V and 1.05\thinspace V a transition occurs which results in current conduction becoming dependent on solution concentration.
Above 1.05\thinspace V the traces diverge showing clear dependence upon concentration.
The point at which this transition occurs is determined by the concentration of saline, with lower concentrations transitioning earlier. A concentration of 0.0625X PBS causes a transition at around 0.95\thinspace V whereas electrodes in the 1.0X PBS solution transition somewhere between 1.0\thinspace and 1.05\thinspace V. The effect of transition can be seen in the green trace at 1.0\thinspace V where the error bar is wide due to the current shift being captured in the measurement at that voltage. Part of a transition with respect to time is visible in figure~\ref{fig:StepResponse_Faradaic} where the red trace drops uncharacteristically after the increment at around 600 seconds.

@@

We attribute the change in behaviour between 0.9\thinspace V and 1.05\thinspace V to a transition to diffusion-controlled conduction.
We hypothesise that the charging of the CPE draws available ions to the electrode, creating a layer of high ionic concentration at the surface irrespective of that of the solution bulk. It is this layer that is subsequently consumed by the Faradaic reactions at a {\color{blue} rate that increases exponentially with} electrode overpotential.  The effect of bulk solution concentration while this layer has formed is negligible until the point at which the layer is consumed faster than it can be replenished. At this stage, and with increasing overpotential, the Faradaic conduction is governed by diffusion of ions from the bulk into the layer, the rate of which increases with the ion concentration in the bulk. We believe this explains the divergence of conduction with concentration between 0.9\thinspace V and 1.05\thinspace V and why there is no observable dependence on bulk ion concentration beforehand.
For the purposes of our model, we are content with placing a limit of 0.9\thinspace V across the pair of interfaces and using the assumption that Faradaic conduction does not change with concentration.

>>>>

\begin{abstract}
We report how the impedance presented by a platinum electrode scales with the concentration of phosphate buffered saline (PBS).
{
    \color{blue}
    We measure the response in various dilutions of PBS with an electrode array as is commonly used in spinal cord stimulator (SCS) implants. We match the parameters of a non-linear electrode-electrolyte interface model to these measurements.
}
We find that the constant phase element of the model scales with approximately the log of concentration, whereas the resistivity is inversely proportional.
Using a novel DC measurement technique we show that the {\color{blue} onset of Faradaic conduction for} a platinum electrode, and thus the safe exposure limit, does not scale with concentration.
We compare objective measurements made in saline to those made in the spinal cavity of live sheep. We comment upon the appropriateness of using PBS as a substitute for in-vivo measurements.
\end{abstract}

\begin{IEEEkeywords}
    Bioelectric phenomena, Bioimpedance, Biomedical electrodes, Biomedical measurements, Biophysics, Electrical stimulation, Implantable biomedical devices
\end{IEEEkeywords}

@@

Having a compact model of an electrode in the appropriate electrolyte allows circuit designers to simulate their designs with valid loads. Knowing the effect upon the model of changes in the electrode or electrolyte would allow designers to anticipate circumstances that will affect the load seen by their circuits. Another appeal of a compact model stems from the fact that electrode characteristics are succinctly and objectively represented by the model parameters. These parameters enable direct comparison of different electrodes, or the change in electrode properties over time. For example, in \cite{Kane13} changes in chronically-implanted electrodes are observed but there is no standard quantitative way of presenting the changes. Similarly, while there is a good understanding of what represents safe exposure, a compact model with parameters permits the prediction of the safety of any given stimulus regime.~\cite{Merrill05}

@@

Electrodes in the laboratory are typically tested in a saline solution selected to mimic the circumstances in which they will operate when implanted. In an implanted setting there is no reference electrode and many of the well established electrochemical measurement regimes do not apply. Here we measure the behaviour in an electrode-electrolyte situation as it would be seen from an implant device.

The electrode used in this work is a commercial linear array of eight platinum electrodes, called an ``Octrode'', intended for Spinal Cord Stimulator (SCS) implantation.~\cite{StJudeOctrode} A picture of an Octrode is presented in figure~\ref{fig:octrode}. A solution of phosphate buffered saline (PBS) diluted to one-tenth concentration (0.1X) is commonly {\color{blue}used as a crude phantom} in the case of SCS, while full concentration (1.0X) PBS is considered more representative of the situation in blood.

Six solutions ranging from 1.0X to 0.025X the concentration (by mass) of a stock PBS solution have been examined.
The ingredients of that stock solution are given in Table~\ref{tab:PBSrecipe}.

@@

    \begin{figure}
        \begin{center}
            \includegraphics[width=230pt]{graphics/interfaceSchematic_noMemristive}
        \end{center}
        \caption{Electrical schematic of the electrode-interface impedance model used in this work. $D_{a}$ and $D_{b}$ represent diodes; CPE and $R_{s}$ represent the constant phase element and series resistance respectively.}
        \label{fig:schematic}
    \end{figure}

    In this manuscript we use an adapted version of the compact non-linear model presented in \cite{ScottSingle2013}. The model is compact in that it is suitable for {\color{blue} entering into freely available electrical circuit simulation software}. The schematic for the {\color{blue} electrode-electrolyte interface} model used is presented as Figure \ref{fig:schematic}.
}

@@

It has been suggested that the primary coupling mechanism between platinum electrodes and physiological saline is the reversible evolution of gas, probably oxygen, at the platinum surface, which dissolves without bubble formation.\cite{Greatbatch1969} Understanding the exact mechanism of displacement or Faradaic charge transfer at the electrode surface is beyond the scope of this work. Rather, we wish to characterise the {electrode-electrolyte interface \color{blue} using a small} {\color{blue} number of} parameters that can be fed into an electrical model.

{
    \color{blue}
    The remainder of this manuscript is broken into five sections. Section \ref{sect:resistorMesh} details the measurement of the inter-electrode resistances {\color{blue}(arising from the bulk resistance of the solution)} and the generation of a resistor network to replicate these resistances.
    Sections \ref{sect:cpe} and \ref{sect:faradaic} are concerned with the measurement and parameter fitting of displacement currents and Faradaic currents, respectively, across the electrolyte interface. Section \ref{sect:in-vivo} compares the results obtained {\color{blue}for} PBS with measurements {\color{blue} made} in-vivo. Finally, in Section \ref{sect:conclusion} we summarise our findings and give concluding comments.
}

@@

The electrical impedance between two electrodes in an electrolyte arises from two interface impedances in series with the resistance due to the bulk of the electrolyte itself. To quantify the impedance presented by a single interface we first need to determine inter-electrode resistance presented by the electrolyte bulk. {\color{blue} Finding these resistances allows us to quantify the {\color{blue}series resistance of the interface ($R_{S}$)} via subtraction.} As there are always two interface impedances between any two electrodes, inter-electrode resistances cannot be measured in isolation.

In \cite{ScottSingle2013}, a series of transresistance measurements were used in conjunction with a physical description of the electrode geometry to build up a representative network of resistors. That resistor network was defined using five parameters: $R_{eri}$, $R_{sri}$, $R_{li}$, mesh depth, and edge depth. Readers are referred to \cite{ScottSingle2013} for the interpretation of these model parameters. Repeating that work, we created a network of resistors that mimic the resistances due to the solution's bulk conductivity.

The mesh depth and padding values of five columns and three rows, respectively, has been taken directly from \cite{ScottSingle2013}. {\color{blue} These parameters, together with an array of 8 electrodes, leads to a generated network of 205 resistors, the layout of which is presented in figure \ref{fig:mesh}.}

A numerical fit {\color{blue} has been} made to the two independent scaling parameters $R_{eri}$ and $R_{li}$\footnote{The values of $R_{eri}$ and $R_{sri}$ are defined by the lengths of the electrode and inter-electrode spacers respectively as described in \cite{ScottSingle2013}. The two parameters are proportional to each other by the ratio of their lengths. Hence, we define $R_{sri}$ to be 3/4 that of $R_{eri}$ making it a dependant parameter.}.
The resulting fit and parameter values are presented in figure~\ref{fig:transresistance} and Table~\ref{tab:RESparams}.

Measurements of the inter-electrode impedance {\color{blue} were made using} a Tektronix TPS2014 DSO with four fully isolated and floating channels; an Agilent 33220A Function Generator; and a desktop PC {\color{blue} running Python 2.7 for instrumentation control and result processing.} The pH level of the stock solution was measured using a calibrated EDT Instruments BA-350 pH meter. It had a pH of 7.4 and no pH adjustments were made to the derived solutions. {\color{blue} The conductivity of solutions was measured using an EDT Instruments RE 388 Tx conductivity meter.}

{
    \color{blue}
    To be clear, the resistive mesh created in this step models \textit{only} the resistivity of the solution bulk. No contribution from the interface itself appears in these measurements. {\color{blue} This measurement is possible} because voltages are always measured across pairs of non-driven electrodes and the measurements are of suitably high impedance. We determine the resistive component of the interface, $R_{S}$, in the {\color{blue} following} section.
}

@@

\section{Displacement Parameters {\color{blue}/ CPE}}
\label{sect:cpe}
Displacement currents are responsible for capacitive behaviour at interfacial boundaries.
They are brought about by the redistribution of charge in response to applied fields, {\color{blue}e.g.,} a charged electrode will attract or repel $Cl^{-}$ from its surface and reorientate polar molecules in the surrounding solution.\cite{Merrill05} This capacitive behaviour may also be the result of electrode polarisation in the form of reversible Faradaic reactions at the surface of the electrode.
Such reactions involving water and platinum, as identified in \cite{Horch2004,Mohtashami2011,Merrill05}, include:
\begin{eqnarray}
    Pt + H_{2}O &\Leftrightarrow& PtO + 2 H^{+} + 2 e^{-}\\
    PtO + H_{2}O &\Leftrightarrow& PtO_{2} + 2 H^{+} + 2e^{-}\\
    Pt + H^{+} & \Leftrightarrow & Pt-H\\
    Pt + H_{2}O + e^{-} &\Leftrightarrow& Pt-H+OH^{-}
\end{eqnarray}

Displacement currents are capable of transferring charge between electrical and biological systems without damaging electrodes or tissue.\cite{Horch2004}
This behaviour is modelled by way of a constant phase element (CPE), also known as a fractional capacitor.
For simulation purposes this element is realised as an array of R-C branches~\cite{ScottSingle2013,Morrison59,Elwakil10} where the value of each R-C pair is chosen to set the slope of impedance relative to frequency. The placement of R-C pairs within the CPE is determined by Morrison's parameters $m$ and $k$, where $k$ determines the spacing, or density, of R-C branches and $m$ determines the resulting slope. {\color{blue} R-C pairs were generated using a Python script to span 1\thinspace nHz --1\thinspace GHz with a spacing of 3 pairs per decade. The CPE contains 55 branches i.e. 110 {\color{blue}electrical} elements are used to represent the CPE.}

We measure displacement currents by sweeping the frequency of a sinusoidal stimulus across electrodes 2 and 7 of the Octrode while measuring the voltage between electrodes 2 and 3 (numbering shown in figure~\ref{fig:octrode}).
This allows for the measurement of a single interface impedance in series with the resistance presented by the bulk solution.

{
    \color{blue}
    Current density during measurements, calculated using an effective surface area of $14\thinspace mm^{2}$, peaked at $161\thinspace \mu A/cm^{2}$ at 10\thinspace kHz in 1.0X PBS and fell as low as $197\thinspace nA/cm^{2}$ at 50\thinspace mHz in 0.025X PBS.
}Charge injection per phase is shown in figure~\ref{fig:chargeInjectionVsFrequency} for each solution.
The stimulus waveform was set at 300\thinspace mV (peak) at each measurement point. The measurement instruments used are the same as those used to measure the inter-electrode resistance in the previous section.

\begin{figure}
    \begin{center}
        \includegraphics{graphics/chargeInjectionEffectiveVsFrequency_magnitude}
    \end{center}
    \caption{\color{blue} Per phase charge injected during CPE measurement in each concentrations of PBS{\color{blue}, calculated with an effective surface area of 14\thinspace mm$^{2}$.}}
    \label{fig:chargeInjectionVsFrequency}
\end{figure}

Figure~\ref{fig:CPE_Magnitude} shows the magnitude of the measured response {\color{blue}for each concentration of PBS} accompanied by simulation results. Figure~\ref{fig:CPE_Phase} shows the phase response for the same measurements, again with the simulated response at each concentration. The simulated responses are generated using fitted parameter values for {\color{blue} $m$, $k$, $|Z|\thinspace @\thinspace 1Hz$, and $R_S$, which are} presented in Table~\ref{tab:CPEparams}.

@@

Figure~\ref{fig:CPE_Scaling} shows measured data and the corresponding fits for both the CPE magnitude at 0.05\thinspace Hz and $R_{S+bulk}$. Here we express the CPE offset parameter as $|Z|\thinspace @ \thinspace 1Hz$, but the fits used $|Z|\thinspace @ \thinspace 0.05Hz$, as data points at this frequency are outside the transitional zone where the impedance of the CPE and that of $R_{S+bulk}$ overlap.

The data presented in figure~\ref{fig:CPE_Magnitude} shows a large, linear, relative shift of $|Z|$ with salinity at higher frequencies, where the resistive component of the solution dominates.
Conversely, at lower frequencies where $|Z|$ changes with frequency, the relative change is closer to being logarithmic. It is this region that is modelled by the CPE.
This suggests that the CPE does not chiefly rely on the added salt ions, but they do have some effect.
Instead, the constant-phase effect must arise from phenomena not associated with the salt ions.

@@

\begin{figure}
    \begin{center}
        \includegraphics{graphics/scalingFactors_Displacement_IEEE}
    \end{center}
    \caption{Simulated and measured values of CPE parameters versus concentration of PBS. $|Z|$ is compared at 0.05\thinspace Hz as measured data is unaffected by $R_{S+bulk}$ at this frequency.}
    \label{fig:CPE_Scaling}
\end{figure}

\begin{table}
    \caption{Displacement parameter scaling.\hspace{\textwidth}$conc$ refers to the mass concentration of PBS.}
    \label{tab:CPEparams}
    \begin{center}
        \begin{tabular}{r | l}
            Parameter & Value \\
            \hline
            $m$ & $1.34$ \\
            $k$ & $1.773$\\
            $|Z|\: @\: 1Hz$ $(\Omega)$& $3284 \times conc^{-0.158}$ \\
            $R_{S}$ $(\Omega)$& $13.38 \times conc^{-0.8397} $\\
        \end{tabular}
    \end{center}
\end{table}

@@

Our model uses two reverse-connected diodes to represent Faradaic current conduction between electrode and electrolyte. Specifically, we use the diodes to model irreversible Faradaic reactions. (Any reversible reactions, those producing reactants bound to the surface of the electrode, are encapsulated by the displacement branch of the model.)
Possible irreversible Faradaic reactions for platinum electrodes {\color{blue}(each with differing reaction potentials)} in saline are:

@@

{\color{blue} To measure Faradaic conduction we must separate its contribution from any displacement behaviour.} By using a DC measurement technique, as opposed to cyclic measurements, we are able to separate the effects in time.
Like a conventional capacitor, the CPE responds {\color{blue} to a change in voltage by absorbing or ejecting charge, causing a spike in current}. The CPE will draw negligible current once it has settled and in the case of our model we assume the remaining current conduction to be the result of Faradaic processes.

@@

Measurements commenced with a 4.5 litre solution of 1.0X PBS which was progressively diluted in factors of two between measurements. The solution was mixed continuously using a standard laboratory grade magnetic stirrer. The solution was in equilibrium with air prior to and during measurement and the electrodes remained submerged at all times.
A single measurement run involved stepping the voltage across the electrodes from 0.5\thinspace V to 1.2\thinspace V in increments of 0.05\thinspace V; this range was previously determined to capture the onset of Faradaic conduction.

@@

In \cite{Cogan2008} it has been noted that results obtained using cyclic voltammetry are dependant on, among other factors, the measurement sweep rate. This indicates a lack of isolation between displacement and Faradaic mechanisms. Our measurements were repeated using wait-times of 4, 16, 32 and 64 seconds between voltage increments, allowing us to determine the effect of sweep rate. From these measurements and those of a separate investigation\footnote{We took measurements of the interface's response to voltage steps when left to settle for 10 000 seconds, of which the first 250\thinspace seconds are shown in figure~\ref{fig:CPE_currentVsTime}. Those measurements show a trend of decreasing current stabilisation time as the overpotential steps increase. Sixty four seconds after stepping from 0.64\thinspace V to 0.72\thinspace V, the current has stabilised to $95\%$ of its value at 10\,000 seconds. At overpotential steps for higher voltages, the settling time is further reduced. These measurements were made in still solution and we expect stirring will further reduce settling time.} we concluded that a wait-time of 64 seconds is sufficient.

@@

In figure~\ref{fig:StepResponse_Faradaic}, electrical current measurements (solid lines) are plotted against time. Each spike marks the point at which the voltage is incremented by 50\thinspace mV. Note that the amount of charge absorbed by the CPE increases with concentration of PBS, but the current for each concentration converges to approximately the same value. Dotted lines have been added that link the electrical current measurements taken 10 seconds after each increment for 0.125X PBS (lower trace) and 1.0X PBS (upper trace). These dotted traces show results typical of cyclic {\color{blue} measurement} methods, where the electrode overpotential is never constant. This illustrates the benefit of using DC measurements when measuring Faradaic currents.

Figure~\ref{fig:faradaic_logCurrentVsVoltage} shows the settled currents plotted on a log scale, versus the voltage applied across the electrode {\color{blue} pair}. This uses the same data as the previous figure but has been processed so that each point represents the average of the final 24 current measurements at each step. Below 0.9\thinspace V it appears that the concentration of the solution had no identifiable effect on Faradaic conduction as there is overlap between each trace and the mean values are not monotonic with concentration in this region.  Between 0.9\thinspace V and 1.05\thinspace V a transition occurs which results in current conduction becoming dependent on solution concentration.
Above 1.05\thinspace V the traces diverge showing clear dependence upon concentration.
The point at which this transition occurs is determined by the concentration of saline, with lower concentrations transitioning earlier. A concentration of 0.0625X PBS causes a transition at around 0.95\thinspace V whereas electrodes in the 1.0X PBS solution transition somewhere between 1.0\thinspace V and 1.05\thinspace V. The effect of transition can be seen in the green trace at 1.0\thinspace V where the error bar is wide due to the current shift being captured in the measurement at that voltage. Part of a transition with respect to time is visible in figure~\ref{fig:StepResponse_Faradaic} where the red trace drops uncharacteristically after the increment at around 600 seconds.

@@

We attribute the change in behaviour between 0.9\thinspace V and 1.05\thinspace V to a transition to diffusion-controlled conduction.
We hypothesise that the charging of the CPE draws available ions to the electrode, creating a layer of high ionic concentration at the surface irrespective of that of the solution bulk. It is this layer that is subsequently consumed by the Faradaic reactions at a {\color{blue} rate that increases exponentially with} electrode overpotential.  The effect of bulk solution concentration while this layer has formed is negligible until the point at which the layer is consumed faster than it can be replenished. At this stage, and with increasing overpotential, the Faradaic conduction is governed by diffusion of ions from the bulk into the layer, the rate of which increases with the ion concentration in the bulk. We believe this explains the divergence of conduction with concentration between 0.9\thinspace V and 1.05\thinspace V and why there is no observable dependence on bulk ion concentration beforehand.
For the purposes of our model, we are content with placing a limit of 0.9\thinspace V across the pair of interfaces and using the assumption that Faradaic conduction does not change with concentration. {\color{blue} As we are primarily interested with capturing only the onset of Faradaic conduction, this limitation does not degrade the usefulness of the model for implant modelling purposes.}

    Figure~\ref{fig:faradaic_currentVsTime} shows a subset of the data used in figure~\ref{fig:StepResponse_Faradaic} (the 1.0X PBS trace) to show the fit between simulated and measured data in the time domain. Here the simulation contains fitted diode parameters, shown in Table~\ref{tab:FaradaicParams}, which give rise to the exponentially increasing steady state current.
It is apparent from this trace that the temporal response of the CPE does not match the transient decay curves seen in the Faradaic measurements, something the CPE should predict. These transient decay curves appear to have a time-constant\footnote{We use the term ``time-constant'' although a CPE does not have a simple exponential time-domain response.} that decreases with each subsequent measurement. This variable time-constant phenomena is especially clear when looking at figure~\ref{fig:CPE_currentVsTime}. Although the Faradaic response transients appear to approach the time-constant predicted by the CPE, there is still behaviour not completely described by the CPE.
